---
title: "ELO updates"
output: html_notebook
---
Load in matchups
```{r}
pacman::p_load(tidyverse, dplyr, ggthemes, data.table, lubridate, glmnet,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, coefplot, skimr, car, ggrepel, slider, lubridate,
               tidymodels,ranger,vip,ggplot2, tune,dials,pdp, purrr, stringr, lmtest,
               sandwich)
```


```{r}
matchups_all_nested = fread('../data/results/matchups_all_nested.csv')
```


Apply pass block win function based on euclidean distance
```{r}

#function to calculate pass block win: encode rusher wins
safe_eval <- function(rusher, blocker, start, end, pass_time, tracking, game_id, play_id) {
  tryCatch({
    if (is.na(start) || is.na(end) || is.na(pass_time) || !is.data.frame(tracking)) {
      return(tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = NA_real_,
        beat_dist    = NA_real_,
        rusher_won   = 0L
      ))
    }

    window_end <- min(end, pass_time, na.rm = TRUE)

    rusher_x_col <- paste0(rusher, "_x")
    rusher_y_col <- paste0(rusher, "_y")
    blocker_x_col <- paste0(blocker, "_x")
    blocker_y_col <- paste0(blocker, "_y")

    wide_rb <- tracking %>%
      filter(
        time_since_snap >= start,
        time_since_snap <= window_end,
        freeze_frame_player %in% c(rusher, blocker)
      ) %>%
      select(time_since_snap,
             player = freeze_frame_player,
             x = freeze_frame_x,
             y = freeze_frame_y) %>%
      group_by(time_since_snap, player) %>%
      summarise(
        x = mean(x, na.rm = TRUE),
        y = mean(y, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      pivot_wider(
        names_from = player,
        values_from = c(x, y),
        names_glue = "{player}_{.value}"
      ) %>%
      left_join(
        tracking %>% select(time_since_snap, qb_x, qb_y),
        by = "time_since_snap"
      ) %>%
      mutate(across(ends_with("_x"), as.numeric),
             across(ends_with("_y"), as.numeric))

    required_cols <- c(rusher_x_col, rusher_y_col,
                       blocker_x_col, blocker_y_col, "qb_x", "qb_y")

    if (nrow(wide_rb) == 0 || !all(required_cols %in% names(wide_rb))) {
      return(tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = NA_real_,
        beat_dist    = NA_real_,
        rusher_won   = 0L
      ))
    }

    # Distance to QB
    d_r <- sqrt((wide_rb[[rusher_x_col]] - wide_rb$qb_x)^2 +
                (wide_rb[[rusher_y_col]] - wide_rb$qb_y)^2)
    d_b <- sqrt((wide_rb[[blocker_x_col]] - wide_rb$qb_x)^2 +
                (wide_rb[[blocker_y_col]] - wide_rb$qb_y)^2)

    idx <- which(d_r < d_b)

    if (length(idx) > 0 && idx[1] <= nrow(wide_rb)) {
      tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = as.numeric(wide_rb$time_since_snap[idx[1]]),
        beat_dist    = as.numeric(d_r[idx[1]]),
        rusher_won   = 1L
      )
    } else {
      tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = NA_real_,
        beat_dist    = NA_real_,
        rusher_won   = 0L
      )
    }

  }, error = function(e) {
    tibble(
      game_id = game_id,
      play_id = play_id,
      rusher_name  = rusher,
      blocker_name = blocker,
      beat_time    = NA_real_,
      beat_dist    = NA_real_,
      rusher_won   = 0L
    )
  })
}

#apply function to get table of results
results2 <- pmap_dfr(
  list(
    rusher     = matchups_all_nested$rusher_name,
    blocker    = matchups_all_nested$blocker_name,
    start      = matchups_all_nested$Engage_start,
    end        = matchups_all_nested$Engage_end,
    pass_time  = matchups_all_nested$pass_time,
    tracking   = matchups_all_nested$data,
    game_id = matchups_all_nested$game_id, 
    play_id = matchups_all_nested$play_uuid
  ),
  safe_eval
)
```

```{r}
# memory cleanup: remove nested data after evaluation
rm(matchups_all_nested)
gc()
```

Cleaning
```{r}
results_cleaned <- results2 %>% 
  drop_na(game_id, rusher_name, blocker_name) %>% 
  arrange(game_id, play_id)

init_play_ratings <- results_cleaned %>%
  select(rusher_name, blocker_name) %>%
  pivot_longer(everything(), names_to = "type", values_to = "player_name") %>%
  mutate(Player_Elo = ifelse(type == "rusher_name", 827, 1200)) %>%  
  distinct(player_name, .keep_all = TRUE)

elo_vec <- setNames(init_play_ratings$Player_Elo, init_play_ratings$player_name)
```

```{r}
# memory cleanup: remove initial ratings after vector creation
rm(init_play_ratings)
gc()
```

Elo function
```{r}
K <- 32
scale <- 319

elo_step <- function(belo, relo, outcome) {
  # expectation for rusher
  E_r <- 1 / (1 + 10^((belo - relo) / scale))
  E_b <- 1 - E_r
  
  # update each player with K * (S - E) where S is actual score
  new_r <- relo + K * (outcome - E_r)      # rusher update
  new_b <- belo + K * ((1 - outcome) - E_b) # blocker update
  
  c(b = new_b, r = new_r)
}

get_elo <- function(id) {
  id <- as.character(id)  
  if (!id %in% names(elo_vec)) elo_vec[[id]] <- 1000
  return(elo_vec[[id]])
} 

```


```{r}

results_cleaned <- as.data.table(results_cleaned)
interaction_counts <- list()  # Initialize interaction count
elo_history <- results_cleaned %>%
  mutate(
    before_rusher_elo = NA_real_,
    before_blocker_elo = NA_real_,
    after_rusher_elo = NA_real_,
    after_blocker_elo = NA_real_, 
    rusher_n = NA_integer_,
    blocker_n = NA_integer_
  )

for (i in seq_len(nrow(elo_history))) {
  r_id <- elo_history$rusher_name[i]
  b_id <- elo_history$blocker_name[i]
  out  <- elo_history$rusher_won[i]

  if (is.na(r_id) || is.na(b_id) || is.na(out)) next

  r_elo <- get_elo(r_id)
  b_elo <- get_elo(b_id)

  elo_history$before_rusher_elo[i]  <- r_elo
  elo_history$before_blocker_elo[i] <- b_elo
  
  interaction_counts[[r_id]] <- if (is.null(interaction_counts[[r_id]])) 1L else interaction_counts[[r_id]] + 1L
  interaction_counts[[b_id]] <- if (is.null(interaction_counts[[b_id]])) 1L else interaction_counts[[b_id]] + 1L

  elo_history$rusher_n[i]  <- interaction_counts[[r_id]]
  elo_history$blocker_n[i] <- interaction_counts[[b_id]]

  new <- elo_step(b_elo, r_elo, out)

  elo_vec[[r_id]] <- new["r"]
  elo_vec[[b_id]] <- new["b"]

  elo_history$after_rusher_elo[i]  <- new["r"]
  elo_history$after_blocker_elo[i] <- new["b"]
}
```

```{r}
#rm(interaction_counts)
#gc()
```


```{r}
#get results
elo_df <- enframe(elo_vec, name = "player_name", value = "final_elo")

#differentiate between rushers and blockers
player_roles <- results_cleaned %>%
  select(rusher_name, blocker_name) %>%
  pivot_longer(cols = everything(), values_to = "player_name") %>%
  mutate(role = ifelse(name == "rusher_name", "Rusher", "Blocker")) %>%
  distinct(player_name, role)

#join to get elo with annotated rusher/blocker label
elo_labeled <- elo_df %>%
  left_join(player_roles, by = "player_name") %>%
  filter(!is.na(role))

#get top 10
top_rushers <- elo_labeled %>%
  filter(role == "Rusher") %>%
  slice_max(final_elo, n = 10)

top_blockers <- elo_labeled %>%
  filter(role == "Blocker") %>%
  slice_max(final_elo, n = 10)
```

```{r}
#pivot to get elo updates per play 
elo_long <- elo_history %>%
  select(game_id, play_id,
         rusher_name, blocker_name,
         after_rusher_elo, after_blocker_elo,
         rusher_n, blocker_n) %>%
  pivot_longer(
    cols = c(after_rusher_elo, after_blocker_elo),
    names_to = "elo_type",
    values_to = "elo"
  ) %>%
  mutate(
    player_name = ifelse(elo_type == "after_rusher_elo", rusher_name, blocker_name),
    interaction_n = ifelse(elo_type == "after_rusher_elo", rusher_n, blocker_n),
    role = ifelse(elo_type == "after_rusher_elo", "Rusher", "Blocker")
  ) %>%
  select(play_id, game_id, player_name, role, interaction_n, elo) %>%
  filter(!is.na(elo))

#filter for top players
top_players <- elo_long %>%
  group_by(player_name, role) %>%
  summarize(final_elo = max(elo, na.rm = TRUE), .groups = "drop") %>%
  group_by(role) %>%
  slice_max(final_elo, n = 10) %>%
  pull(player_name)

elo_top <- elo_long %>%
  filter(player_name %in% top_players)
```


```{r}
elo_top_rusher <- elo_top %>% filter(role == "Rusher")
elo_top_blocker <- elo_top %>% filter(role == "Blocker")

ggplot(elo_top_rusher, aes(x = as.numeric(interaction_n), y = elo, color = player_name)) +
  geom_line(size = 1) +
  labs(
    title = "Top 10 Rushers: Elo Trajectory",
    x = "Interaction Number",
    y = "Elo Rating",
    color = "Rusher"
  ) +
  theme_minimal()

ggplot(elo_top_blocker, aes(x = as.numeric(interaction_n), y = elo, color = player_name)) +
  geom_line(size = 1) +
  labs(
    title = "Top 10 Blockers: Elo Trajectory",
    x = "Interaction Number",
    y = "Elo Rating",
    color = "Rusher"
  ) +
  theme_minimal()

# Save results to processed data
write_csv(elo_history, "../data/results/elo_history.csv")
write_csv(elo_labeled, "../data/results/player_elo_ratings.csv")
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

