---
title: "Hudl 10 Sample Games ELO"
author: "Victoria"
date: "2025-07-02"
output: html_document
---

```{r Setup, include=FALSE, results='hide', warning=FALSE}
knitr::opts_chunk$set(echo = T, fig.width=8, fig.height=4)
options(scipen = 0, digits = 3)  # controls base R output

# Package setup
if(!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, dplyr, ggthemes, data.table, lubridate, glmnet,
               GGally, RColorBrewer, ggsci, plotROC, usmap,
               plotly, ggpubr, vistime, coefplot, skimr, car, ggrepel, slider, lubridate,parallel, 
               tidymodels,ranger,vip,ggplot2, tune,dials,pdp, purrr, stringr, lmtest, future,
               sandwich)

# boost working memory
mem.maxVSize(81920)

plan(multisession, workers = parallel::detectCores() - 1)
```


Read in data
```{r}
#read in five weeks of data
new_data <- fread("../data/processed/full_hudl_data.csv")
```

```{r}
names(new_data)
head(new_data)
```
Encode sacks
```{r}
print(unique(new_data$event_types))

sacks <- new_data %>%
  filter(str_detect(event_types, "Sack")) %>% 
  select(play_uuid, game_id, event_player_name) %>%
  rename(sack_player = event_player_name) %>% 
  distinct(play_uuid, game_id, sack_player) %>% 
  mutate(sack = 1L)

write.csv(sacks, '../data/processed/sacks.csv', row.names = FALSE)
```


Get quarterbacks
```{r}
qbs <- new_data %>% 
  filter(freeze_frame_position == "Quarterback") %>% 
  distinct(freeze_frame_player)
```

Get hits
```{r}
tackles_qb <- new_data %>%
  filter(event_types == "{Tackle}") %>%
  filter(opponent_player_name %in% qbs$freeze_frame_player) %>% 
  filter(freeze_frame_position == "Quarterback") %>%   # Only QB getting tackled
  select(game_id, play_uuid, time_since_snap, event_player_name,
         event_x, event_y, freeze_frame_x, freeze_frame_y,
         freeze_frame_position, play_start_event, event_types, opponent_player_name)

pass_times <- new_data %>%
  filter(str_detect(event_types, "Pass")) %>%
  group_by(play_uuid) %>%
  summarise(pass_time = min(time_since_snap), .groups = "drop")

tackles_qb <- tackles_qb %>%
  left_join(pass_times, by = "play_uuid") %>%
  anti_join(sacks, by = "play_uuid") %>% 
  filter(is.na(pass_time) | time_since_snap < pass_time) %>% 
  filter(!is.na(time_since_snap)) 

hits <- tackles_qb %>% 
  select(play_uuid, game_id, event_player_name) %>%
  rename(hit_player = event_player_name) %>% 
  distinct(play_uuid, game_id, hit_player) %>% 
  mutate(hit = 1L) 

write.csv(hits, '../data/processed/hits.csv', row.names = FALSE)
```

Look at tackle attempts
```{r}
tackles_attempts <- new_data %>%
  filter(event_types == "{\"\"\"\"Tackle Attempt\"\"\"\"}" ) %>%
  filter(opponent_player_name %in% qbs$freeze_frame_player) %>% 
  filter(freeze_frame_position == "Quarterback") %>%   # Only QB getting tackled
  left_join(pass_times, by = "play_uuid") %>% 
  select(game_id, play_uuid, time_since_snap, pass_time, event_player_name,
         event_x, event_y, freeze_frame_x, freeze_frame_y,
         freeze_frame_position, event_types, opponent_player_name) %>% 
  anti_join(sacks, by = "play_uuid") %>% 
  anti_join(hits, by = "play_uuid") 
```

Filter 
```{r}
matchups <- new_data %>% 
  select(freeze_frame_player, event_player_name, opponent_player_name, event_uuid, event_game_index, game_id, play_uuid, player_id,  time_since_snap,opponent_player_id,  freeze_frame_player_id, event_types, event_x, event_y, start_engagement_uuid, end_engagement_uuid, play_start_event, time_since_snap, freeze_frame_x, freeze_frame_y, event_player_position, freeze_frame_player_id, freeze_frame_position, freeze_frame_player_team, nflfast_game_id)

```

Add the freeze frame of opponent players and qb in matchups to data 
```{r}

matchups <- new_data

matchup_tracking <- matchups %>%
  left_join(
    matchups %>% 
      select(game_id, play_uuid, time_since_snap, freeze_frame_player_id, freeze_frame_x, freeze_frame_y, freeze_frame_player, event_game_index) %>% 
      rename(
        opponent_player_id = freeze_frame_player_id,
        x_opp = freeze_frame_x,
        y_opp = freeze_frame_y,
        opponent_name = freeze_frame_player
      ),
    by = c("game_id", "play_uuid", "time_since_snap", "opponent_player_id")
  )

matchup_tracking_with_qb <- matchups %>%
  left_join(
    matchups %>%
      filter(freeze_frame_position == "Quarterback") %>%
      # %>%
      select(game_id, play_uuid, time_since_snap,
             qb_id = freeze_frame_player_id,
             qb_x  = freeze_frame_x,
             qb_y  = freeze_frame_y,
             qb_name = freeze_frame_player),
    by = c("game_id", "play_uuid", "time_since_snap")
  )

```

Make table with engagement start and end times
```{r}
engagements <- matchup_tracking_with_qb %>% 
  filter(event_types %in% c("{\"\"\"\"Engagement Start\"\"\"\"}"  , "{\"\"\"\"Engagement End\"\"\"\"}", "{Pass}" , "{Pass,Hit}", "{Hit,Pass}", "{Sack,Tackle}"))

```

Identify rusher and blocker and put engagement start and end in columns
```{r}
#make table with each 
matchups_all <- engagements %>%
  filter(
    event_types %in% c(
      "{\"\"\"\"Engagement Start\"\"\"\"}",
      "{\"\"\"\"Engagement End\"\"\"\"}",
      "{Pass}", "{Pass,Hit}", "{Sack,Tackle}", "{Hit,Pass}"
    )
  ) %>%
  arrange(play_uuid, event_game_index, player_id, opponent_player_id, event_types) %>%
  #filter(player_position %in% c("Defensive Lineman", "Linebacker")) %>%  # only keep defenders
  group_by(play_uuid, player_id, opponent_player_id) %>%
  summarize(
  rusher_idx = which(freeze_frame_position %in% c("Defensive Lineman", "Linebacker"))[1],
  blocker_idx = which(opponent_position == "Offensive Lineman")[1],

  rusher_id = if (!is.na(rusher_idx)) player_id[rusher_idx] else NA_integer_,
  rusher_name = if (!is.na(rusher_idx)) freeze_frame_player[rusher_idx] else NA_character_,

  blocker_id = if (!is.na(blocker_idx)) opponent_player_id[blocker_idx] else NA_integer_,
  blocker_name = if (!is.na(blocker_idx)) opponent_player_name[blocker_idx] else NA_character_,

  Engage_start = time_since_snap[event_types == "{\"\"\"\"Engagement Start\"\"\"\"}"][1],
  Engage_end   = time_since_snap[event_types == "{\"\"\"\"Engagement End\"\"\"\"}"][1],

  game_id  = first(game_id),
  qb_id    = first(qb_id),
  qb_x     = first(qb_x),
  qb_y     = first(qb_y),
  qb_name  = first(qb_name),
  .groups = "drop"
)

matchups_all <- matchups_all %>%
  drop_na(blocker_idx) %>%
```
Clear memory
```{r}
rm(matchups, matchup_tracking, matchup_tracking_with_qb)
gc()
```


Find pass times and add to table
```{r}
#get pass times
pass_times <- engagements %>%
  filter(event_types == "{Pass}") %>%
  group_by(play_uuid) %>%
  summarize(pass_time = min(time_since_snap, na.rm = TRUE), .groups = "drop")

matchups_all <- matchups_all %>%
  left_join(pass_times, by = "play_uuid")
```
Add a column to each engagement with each play nested in that row's entry
```{r}
#prepare engagements table to iterate through frames in a play
play_tracking <- engagements %>%
  group_by(play_uuid) %>%
  nest()

```


```{r}
#merge with matchups_all to contain each matchups and then the freeze frames for each one in a separate column
matchups_all_nested <- matchups_all %>%
  left_join(play_tracking, by = "play_uuid")

```

```{r}

#function to calculate pass block win: encode rusher wins
safe_eval <- function(rusher, blocker, start, end, pass_time, tracking, game_id, play_id) {
  tryCatch({
    if (is.na(start) || is.na(end) || is.na(pass_time) || !is.data.frame(tracking)) {
      return(tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = NA_real_,
        beat_dist    = NA_real_,
        rusher_won   = 0L
      ))
    }

    window_end <- min(end, pass_time, na.rm = TRUE)

    rusher_x_col <- paste0(rusher, "_x")
    rusher_y_col <- paste0(rusher, "_y")
    blocker_x_col <- paste0(blocker, "_x")
    blocker_y_col <- paste0(blocker, "_y")

    wide_rb <- tracking %>%
      filter(
        time_since_snap >= start,
        time_since_snap <= window_end,
        freeze_frame_player %in% c(rusher, blocker)
      ) %>%
      select(time_since_snap,
             player = freeze_frame_player,
             x = freeze_frame_x,
             y = freeze_frame_y) %>%
      group_by(time_since_snap, player) %>%
      summarise(
        x = mean(x, na.rm = TRUE),
        y = mean(y, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      pivot_wider(
        names_from = player,
        values_from = c(x, y),
        names_glue = "{player}_{.value}"
      ) %>%
      left_join(
        tracking %>% select(time_since_snap, qb_x, qb_y),
        by = "time_since_snap"
      ) %>%
      mutate(across(ends_with("_x"), as.numeric),
             across(ends_with("_y"), as.numeric))

    required_cols <- c(rusher_x_col, rusher_y_col,
                       blocker_x_col, blocker_y_col, "qb_x", "qb_y")

    if (nrow(wide_rb) == 0 || !all(required_cols %in% names(wide_rb))) {
      return(tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = NA_real_,
        beat_dist    = NA_real_,
        rusher_won   = 0L
      ))
    }

    # Distance to QB
    d_r <- sqrt((wide_rb[[rusher_x_col]] - wide_rb$qb_x)^2 +
                (wide_rb[[rusher_y_col]] - wide_rb$qb_y)^2)
    d_b <- sqrt((wide_rb[[blocker_x_col]] - wide_rb$qb_x)^2 +
                (wide_rb[[blocker_y_col]] - wide_rb$qb_y)^2)

    idx <- which(d_r < d_b)

    if (length(idx) > 0 && idx[1] <= nrow(wide_rb)) {
      tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = as.numeric(wide_rb$time_since_snap[idx[1]]),
        beat_dist    = as.numeric(d_r[idx[1]]),
        rusher_won   = 1L
      )
    } else {
      tibble(
        game_id = game_id,
        play_id = play_id,
        rusher_name  = rusher,
        blocker_name = blocker,
        beat_time    = NA_real_,
        beat_dist    = NA_real_,
        rusher_won   = 0L
      )
    }

  }, error = function(e) {
    tibble(
      game_id = game_id,
      play_id = play_id,
      rusher_name  = rusher,
      blocker_name = blocker,
      beat_time    = NA_real_,
      beat_dist    = NA_real_,
      rusher_won   = 0L
    )
  })
}

#apply function to get table of results
cat("Starting parallel evaluation of", nrow(matchups_all_nested), "engagements...\n")
results2 <- future_pmap_dfr(
  list(
    rusher     = matchups_all_nested$rusher_name,
    blocker    = matchups_all_nested$blocker_name,
    start      = matchups_all_nested$Engage_start,
    end        = matchups_all_nested$Engage_end,
    pass_time  = matchups_all_nested$pass_time,
    tracking   = matchups_all_nested$data,
    game_id    = matchups_all_nested$game_id, 
    play_id    = matchups_all_nested$play_uuid
  ),
  safe_eval,
  .options = furrr_options(seed = TRUE)
)

cat("Parallel evaluation completed. Processing", nrow(results2), "results.\n")
```

```{r}
write.csv(results2, '../data/processed/results2.csv')
```


